Refinement of Thermostated Molecular Dynamics via Backward Error Analysis

### 2018-09-17 rigid_bodies.tex
Some small changes

This paper proceeds as follows.
In Secs.~\ref{sec:nve}-\ref{sec:nvt} we review the rigid-body dynamics in the NVE and NVT ensembles.
This is followed by our developments on backward error analysis, which we employ in Sec.~\ref{sec:refined_method} to derive our new NVT scheme, whose performance we assess in Sec.~\ref{sec:numerical_results}.
Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}. Appendix \ref{sec:rigid body shadow hamiltonian} contains a detailed derivation of the shadow Hamiltonian approximation.

@@

As an example of application in stochastic thermostats, we consider the  velocity-rescaling scheme introduced by Bussi \textit{et al}. \cite{Bussi_2007}.
Its refined version is simply obtained by substituting $K$ by $\refined{K}$ in Eq.~(A7) of Ref.~\citenum{Bussi_2007}.

@@

\subsection{Correction of Discretization Effects on Thermodynamic Properties}

@@

In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid-water system using several NVT-MD methods with various time-step sizes.
All simulations were carried out for $T = 298.0~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
the mean potential energy per molecule $\langle U/N \rangle$,
the mean internal virial per molecule $\langle W/N \rangle$, and
the constant-volume specific heat $c_\mini{V} = \frac{\langle \Ham{}^2 \rangle - \langle \Ham{} \rangle^2}{N k_\mini{B} T^2}$.
The internal virial $W$ is computed as described in Ref.~\citenum{Silveira_2017}.
All symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
The horizontal dotted lines are meant to help visualize the departure of the computed ensemble averages from their expected values, which correspond to average results obtained with $\timestep = 1~\text{fs}$ by employing the three refined methods (details below).
Error bars represent 95\% confidence intervals obtained by using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.

@@

Fig.~\ref{fig:properties} contains results of five different NVT integrators.
Our measure-preserving extension to rigid bodies of the splitting scheme by Martyna \textit{et al}. \cite{Martyna_1996} is the basis of three of them.
Recall that this is a deterministic method in which a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} is attached to all degrees of freedom.
First, the method is applied in a conventional, refinement-free way, with the unsplit solution of Ref.~\citenum{Silveira_2017} being used for handling rotations.
Then, it is modified by introducing the refinement/reweighting procedure proposed in Sec.~\ref{sec:refined_method}.
In Fig.~\ref{fig:properties}, these two integrators are referred to as ``Martyna'' and ``Refined/NHC'', respectively.
Next, we replace the unsplit rotation solution by the simpler NO-SQUISH splitting of Miller \textit{et al}. \cite{Miller_2002}, denoting the resulting integrator by ``Refined/Miller'' in Fig.~\ref{fig:properties}.
We note that our refinement procedure is rigorously valid (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations only.
Nevertheless, it is known that NO-SQUISH rotations approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
The fourth integrator we test is a refined version of the velocity-rescaling method of Bussi \textit{et al}. \cite{Bussi_2007}, which differs from the others for involving a single stochastic thermostat.
This is identified in Fig.~\ref{fig:properties} as ``Refined/Bussi''.
Finally, we also consider the double-thermostat integrator of Kamberaj \textit{et al}. \cite{Kamberaj_2005} for comparison purposes, despite the spurious energy flux between thermostats described in the preceding section.
This is done to show that the method, denoted by ``Kamberaj'' in Fig.~\ref{fig:properties}, also exhibits discretization-related issues not present in other integrators.

Before discussing the results shown in Fig.~\ref{fig:properties}, we stress that the thermodynamic properties are estimated in distinct ways for conventional and refined integrators.
For the conventional ones, ``Martyna'' and ``Kamberaj'', simple averages are taken from the sampled configurations.
For the refined ones, reweighting is carried out via Eq.~\eqref{eq:reweighting}.

In Fig.~\ref{fig:properties}(a), we see that the estimated temperature for ``Kamberaj'' decreases appreciably with increasing $\timestep$.
This behavior is typical of Verlet-like NVE integrators\cite{Davidchack_2010, Silveira_2017}, reflecting that particle momenta is weakly modified by the NHC factorization scheme of Kamberaj \textit{et al}. \cite{Kamberaj_2005}.
On the other hand, the stringent temperature control achieved in ``Martyna'' leads to a severely degraded accuracy in both $\langle U \rangle$ and $\langle W/N \rangle$, as noted in Figs.~\ref{fig:properties}(b)-(c).
According to the results of Fig.~\ref{fig:nve}, discretization errors in the NVE integration become manifest at $2~\text{fs}$.
Thus, the apparent success of the scheme by Martyna \textit{et al.} \cite{Martyna_1996}, in terms of the kinetic energy control for the entire range of time-step sizes considered, is wiped out by those discretization errors which, if not properly handled, yield considerably degraded configurational properties.

>>>>

This paper proceeds as follows.
In Secs.~\ref{sec:nve} and \ref{sec:nvt} we review the rigid-body dynamics in the NVE and NVT ensembles.
This is followed by our developments on backward error analysis, which we employ in Sec.~\ref{sec:refined_method} to derive a new NVT scheme, whose performance we assess in Sec.~\ref{sec:numerical_results}.
Finally, we present some concluding remarks in Sec.~\ref{sec:conclusion}. Appendix \ref{sec:rigid body shadow hamiltonian} contains a detailed derivation of the shadow Hamiltonian approximation for a system containing rigid bodies.

@@

As an example of application in stochastic thermostats, we consider the  velocity-rescaling scheme introduced by Bussi \textit{et al}. \cite{Bussi_2007}.
Its refined version can be simply obtained by substituting $K$ by $\refined{K}$ in Eq.~(A7) of Ref.~\citenum{Bussi_2007}.
In the implementation used to obtain some of the results in Sec.~\ref{sec:discretization effects}, we also split the stochastic propagator in order to have a Trotter-Suzuki type of integration (whereas a leap-frog type is used originally).

@@

\subsection{Correction of Discretization Effects on Thermodynamic Properties}
\label{sec:discretization effects}

@@

In Fig.~\ref{fig:properties}, we present some thermodynamic properties computed for our liquid-water system using several NVT-MD methods with various time-step sizes.
All simulations were carried out for $T = 298~\text{K}$ and the analyzed quantities are the estimated temperature $T_\text{est} = \frac{2\langle K \rangle}{(6N-3)k_\mini{B}}$,
the mean potential energy per molecule $\langle U/N \rangle$,
the mean internal virial per molecule $\langle W/N \rangle$, and
the constant-volume specific heat $c_\mini{V} = \frac{\langle \Ham{}^2 \rangle - \langle \Ham{} \rangle^2}{N k_\mini{B} T^2}$.
The internal virial $W$ is computed as described in Ref.~\citenum{Silveira_2017}.
All symbols in Fig.~\ref{fig:properties} represent results obtained after $18~\text{ns}$ of production time, while the lines are just guides for the eyes.
The horizontal dotted lines are meant to help visualize the departure of the computed properties from their expected values, which correspond to average results obtained with $\timestep = 1~\text{fs}$ for the three refined methods (details below).
Error bars represent 95\% confidence intervals obtained by using the Overlapping Batch Means estimator \cite{Meketon_1984, Flegal_2010} and standard uncertainty propagation formulas.

@@

Fig.~\ref{fig:properties} contains results of five different NVT integrators.
Our measure-preserving extension to rigid bodies of the splitting scheme by Martyna \textit{et al}. \cite{Martyna_1996} is the basis of three of them.
Recall that this is a deterministic method in which a single Nos\'{e}-Hoover chain thermostat \cite{Martyna_1992} is attached to all degrees of freedom.
First, the method is applied in a conventional, refinement-free way, with the unsplit solution of Ref.~\citenum{Silveira_2017} being used for handling rotations.
Then, it is modified by introducing the refinement/reweighting procedure proposed here in Sec.~\ref{sec:refined_method}.
In Fig.~\ref{fig:properties}, these two integrators are referred to as ``Martyna'' and ``Refined/NHC'', respectively.
Next, we replace the unsplit rotation solution by the simpler NO-SQUISH splitting of Miller \textit{et al}. \cite{Miller_2002}, denoting the resulting integrator by ``Refined/Miller'' in Fig.~\ref{fig:properties}.
We note that our refinement procedure is only rigorously valid (see Appendix \ref{sec:rigid body shadow hamiltonian}) for unsplit rotations.
Nevertheless, it is known that NO-SQUISH rotations approach them almost indistinguishably in the case of water if small time steps are employed \cite{Silveira_2017}.
The fourth integrator we test is a refined version of the velocity-rescaling method of Bussi \textit{et al}. \cite{Bussi_2007}, which differs from the others for involving a global stochastic thermostat.
This is identified in Fig.~\ref{fig:properties} as ``Refined/Bussi''.
Finally, we also consider the double-thermostat integrator of Kamberaj \textit{et al}. \cite{Kamberaj_2005} for comparison purposes, despite its spurious energy flux described in the preceding section.
This is done to show that the method, denoted by ``Kamberaj'' in Fig.~\ref{fig:properties}, also exhibits discretization-related issues not present in other integrators.
We stress that thermodynamic properties are estimated in distinct ways for conventional and refined integrators.
For the former, simple averages are taken from the sampled configurations.
For the latter, reweighting is carried out via Eq.~\eqref{eq:reweighting}.

In Fig.~\ref{fig:properties}(a), we see that the estimated temperature for the thermostat identified as ``Kamberaj'' decreases appreciably as the time step size increases.
This behavior is typical of Verlet-like NVE integrators\cite{Davidchack_2010, Silveira_2017}, indicating that the effect of the NHC factorization scheme of Kamberaj \textit{et al}. \cite{Kamberaj_2005} on the particle momenta is weak.
On the other hand, the stringent temperature control achieved by the thermostat named ``Martyna'' leads to a severely degraded accuracy in both $\langle U/N \rangle$ and $\langle W/N \rangle$, as noted in Figs.~\ref{fig:properties}(b) and \ref{fig:properties}(c).
According to the results of Fig.~\ref{fig:nve}, discretization errors in the NVE integration become manifest at $2~\text{fs}$.
Thus, the apparent success of the scheme by Martyna \textit{et al.} \cite{Martyna_1996}, in terms of the kinetic energy control for the entire range of time-step sizes considered, is wiped out by those discretization errors which, if not properly handled, yield considerably degraded configurational properties.
